\documentclass[a4paper,11pt]{article}

% Page geometry
\usepackage[margin=2cm,top=2.5cm,bottom=2.5cm]{geometry}

% Essential packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype} % Better typography

% COLORS (matching your charts: warm beige + teal accent)
\usepackage[dvipsnames]{xcolor}
\definecolor{background}{HTML}{f8fafc}
\definecolor{surface}{HTML}{f8fafc}
\definecolor{textmuted}{HTML}{7a7974}
\definecolor{accent}{HTML}{01696f}
\definecolor{grid}{HTML}{dcd9d5}
\definecolor{success}{HTML}{437a22}
\definecolor{deepgreen}{HTML}{2e5c10}

% Modern font (Satoshi/General Sans equivalent)
\usepackage[scaled]{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{libertinus}


% Tables
\usepackage{booktabs}
\usepackage{siunitx} % Numbers

% Listings (code)
\usepackage{listings}
\usepackage{xcolor}
\lstset{
  basicstyle=\ttfamily\small\color{black},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{background!90},
  numbers=left,
  numberstyle=\tiny\color{textmuted},
  keywordstyle=\color{accent},
  commentstyle=\color{black},
  stringstyle=\color{success}
}

% Bibliography
\usepackage[style=numeric,autocite=superscript,backend=biber,sorting=none]{biblatex}
\addbibresource{./report.bib}

% Graphics and links
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=accent,
  urlcolor=accent,
  citecolor=accent,
  pdfborder={0 0 0}
}

% Headers/footers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small Python Data Visualization Report}
\fancyhead[R]{\thepage}
\fancyfoot[C]{\small Itziar Morales Rodríguez}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\headrule}{\color{grid}\hrule height 0.4pt}

% Title formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries\color{accent}}{\thesection}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{black}}{\thesubsection}{0.5em}{}
\titlespacing*{\section}{0pt}{1.5em}{0.8em}
\titlespacing*{\subsection}{0pt}{1.2em}{0.6em}

% Custom colors for enumerations
\usepackage{enumitem}
\setlist[itemize]{leftmargin=*}
\setlist[enumerate]{leftmargin=*}

% Background (subtle)
\usepackage{afterpage}
\AfterPage{\pagecolor{background}\color{black}}

\title{\color{accent}\Huge Python Data Visualization\\Assignment Report}
\author{\Large\color{black}Itziar Morales Rodríguez}
\date{\large\color{black}April 12, 2026}

\begin{document}
\pagecolor{background}\color{black}

\maketitle
\vspace{1em}
\hrule height 1pt \color{black}

% Table of Contents
\tableofcontents
\newpage

\section{Introduction}

\par
The aim of this project is to explore data visualization using Python, from
selecting the data all the way to telling a story through it's visualization.

\subsection{Methodology}

\par
The project itself followed the following process to achieve the previously
mentioned goal:
\begin{enumerate}
	\item Dataset selection
	\item Data analysis and processing
	\item Questions derived from the dataset
	\item Visualization types needed to answer each question and storytell
	\item Further data processing needed for each visualization
	\item Library selection for each visualization
	\item Package visualizations within a combined story narrative that tells a
	      story about the data
\end{enumerate}

\subsection{Sources}

\par
Numerous sources are used through out the project such as:

\begin{itemize}
	\item AI
	\item Documentation
	\item Forums
	\item Literature
	\item Lecture notebooks
\end{itemize}

\par
Regardless of the source, these are indicated through bibliography/references
or in the case of AI, an explicit note indicating the model and prompt used.

\subsection{AI Usage}

\par
A point was made to use AI in this project in a manner in which it's answers
are scrutinized and reviewed, to amke sure hallucinations are not taken into
account and answers are not just implemented without a second thought. This
way, an activate usage of AI is incentivized with certain restraints in place.

\par
AI being used for this project was also hosted locally, using smaller models to
do so to encourage reproduciability (since prompts are also included within
the notebook and report).

\section{Dataset}

\par
Data for this project was required to not have been selected by a different
student. To minimize the risk of this happening, activity from a Gamrmin
Forerunner 245 was chosen, reducing the chances of another student having the
same dataset. The data located in \texttt{./data/activities.csv} consists of:

\begin{itemize}
	\item \textbf{Shape}: 31 columns \& 180 rows
	\item \textbf{Date span}: 21 February 2021 - 30 March 2026 (5 years)
	\item \textbf{Data types}: Dates, numerics, categorical, geospatial (Title
	      contains locations)
	\item \textbf{Key columns}: Date, Activity Type, Time, Distance, Avg HR
	\item \textbf{Activity types}: Running, Walking, Cycling, Strength, Swimming,
	      etc.
\end{itemize}

\par
Having 31 columns allows for many different questions to be asked about the
dataset, specially since some of these columns allow for other data fields to
be extracted (activity intensity from Avg HR). The limited number of rows (180)
whilst still being a decent amount, allows it's content to be analyzed in a
more hands on way, being able to be more critical about each specific row and
actively engage in processing in a more granular fashion. The dates spanning 5
years back allows for a story to be told through time, and analyze factors
such as consistency. Lastly, varying data types allows for distinct types of
visualizations that require specific untis to be used such as coordinates
(from location).

\subsection{Analysis}

\par
Before asking questions about the dataset, it was further analyzed using both
the \texttt{df.info()} function to check how the data was imported into the
pandas dataframe. A small AI analysis was invoked using qwen3.5:9b for the
prompt which was the following:

\begin{center}
	Read the linked CSV file and analyze it, create a table with the following
	fields:
	\begin{itemize}
		\item N: Column number
		\item Column: Column name
		\item Description: Content description (what the data in the column usually)
		      looks like), excluding null indicators
		\item Null indicator: Indicator that seems to be used for when the data is
		      null
		\item Data type: Suggested data type to be used when reading CSV in Pandas.
	\end{itemize}
\end{center}

\par
The table generated was then used as a starting point although it's content was
scrutinized and interprerted correctly, assessing whether or not the AI had
hallucinated when generating it. Some important conclusions reached with the
help of this analysis (done in section 2.1 of \texttt{data\_viz\_notebook.ipynb})
where:

\begin{itemize}
	\item Some columns weren't really needed or used, such as \textit{Favorite}
	\item "\texttt{--}" was used as a null indicator on most columns
	\item Correct data types for each column
	\item Some columns belonged to specific activity types, such as \textit{Avg.
		      Swolf} for swimming
\end{itemize}

\subsection{Processing}

\par
Key processing steps (section 3 of \texttt{data\_viz\_notebook.ipynb}):

\begin{itemize}
	\item Replaced `--` nulls with NaN
	\item Converted Time/Distance to numeric (regex parsing)
	\item Extracted locations from Title
	\item Created WHO-equivalent activity minutes\autocite[32]{WHO2020} from
	      Avg HR thresholds
	\item Weekly aggregation for time series analysis
	\item Activity Type simplification (treadmill running, trail running,
	      running are all running)
\end{itemize}

Libraries: Pandas\autocite{pandas}, NumPy\autocite{numpy}, re\autocite{re}
(regex)

Some of the data processing was effectuated in order to satisfy the data
requirements of some of the visualizations, which will be explained in the
following section.

\subsection{Research Questions}

Three questions were formulated based on data exploration:

\begin{enumerate}
	\item What types of activities dominate the dataset and their frequencies?
	\item How do activity patterns vary by location?
	\item Do recorded activities show case "enough" weekly exercise?
\end{enumerate}

\subsubsection{Question 1: Activity Composition}\label{q1}

\par
Question 1 is a good starting point because it allows for the superficial
exploratory analysis of the data. This facilitates the understanding of what
activity types end up shaping most of the data, and is a good starting point
that is taken into account when further exploring other questions.

\subsubsection{Question 2: Geographic Patterns}\label{q2}

\par
Question 2 proceeds naturally from question 1, as it further dives into
understanding the nature of the data, in this case, classifying it into
geogrpahic location rather than categorical activity types. Furthermore, it
also enables the use of different visualization methods due to the unique
geographic data type, location.

\subsubsection{Question 3: Weekly Activity Assessment}\label{q3}

\par
Question 3 is purposefully subjective to some degree, to allow for a variety of
approaches to visualizing and analysing the data, alongside more liberty when
it comes to choosing what narrative to pursue in storytelling. This allows for
many different charts to be used, and for data to be used in creative ways.

\section{Storytelling}

\par
The focus in Storytelling will be the following: A fitness conscious user who
has a garmin watch, has measured their activities multiple years. With
questions 1 and 2 we can come to understand whether the user participates in
varied activity types regardless of their location (suggesting that they might
sustain their fitness throughout holidays). Through the third question, we can
come to understand whether this data being recorded, suggests that the user is
engaging in enough fitness activity to be considered healthy, what periods has
this happened in, and is the data trending in a positive direction.

\par
We get to explore the fitness journey of an individual through visualizations,
leaving us with a deep understanding of what their situation is, what is good
about their current fitness, and how they've improved through out time.

\section{Visualization Summary}

\begin{table}[h]
	\centering
	\begin{tabular}{p{2cm}p{3cm}p{3cm}p{3cm}}
		\toprule
		Question & Charts & Chart Type     & Library    \\
		\midrule
		Q1       & Fig.   & Horizontal bar & Seaborn    \\
		         &        &                & Matplotlib \\
		Q2       & Fig.   & Map            & Plotly     \\
		         &        &                & Bokeh      \\
		Q3       & Fig.   & Heatmap        & Seaborn    \\
		         &        & Timeline       & Matplotlib \\
		         & Fig.   & Lollipop       & ipywidgets \\
		         &        & (interactive)  & Matplotlib \\
		         & Fig.   & Scatter+trend  & Seaborn    \\
		         &        &                & Matplotlib \\
		\bottomrule
	\end{tabular}
	\caption{Visualization summary meeting all requirements (3+ questions, 5+ charts, 3 libraries, 1+ data source).}
	\label{tab:summary}
\end{table}

\section{Library Choices \& Best Practices}

\textbf{Matplotlib}: Core plotting, manual layouts, custom colormaps. Used for precise control (heatmaps, timelines).

\textbf{Seaborn}: Statistical plots (barplot, regplot). Simplified complex visualizations while maintaining customization.

\textbf{ipywidgets}: Interactivity for year comparison. Enhanced exploratory analysis.

\textbf{Best practices applied} (Unit 1):
\begin{itemize}
	\item Color encoding follows perceptual principles (sequential for magnitude)
	\item Minimal ink: removed grids, tight spacing
	\item Clear hierarchy: prominent thresholds, trend lines
	\item Accessibility: high contrast, clear typography
\end{itemize}

\section{Conclusions}

The analysis reveals:
\begin{itemize}
	\item Running dominates training (47\%) but consistency improved 2025–2026
	\item WHO targets met sporadically; 2026 shows upward trend but needs sustained 300+ min weeks
	\item Location patterns suggest trail runs yield higher intensity
\end{itemize}

Future work: integrate weather data via API, add interactive dashboard with Plotly Dash.

\printbibliography

\end{document}
